Automated Assessment of Residual Plots with Computer Vision Models

### 2025-06-16 paper.Rmd
Adjust plot size

```{r false-finding, fig.pos="!h", fig.width = 4, fig.height = 4*4/5, fig.pos="!h", fig.cap="An example residual vs fitted values plot (red line indicates 0 corresponds to the x-intercept, i.e. $y=0$). The vertical spread of the data points varies with the fitted values. This often indicates the existence of heteroskedasticity, however, here the result is due to skewed distribution of the predictors rather than heteroskedasticity. The Breusch-Pagan test rejects this residual plot at 95\\% significance level ($p\\text{-value} = 0.046$)."}
set.seed(452)
ori_x <- rand_lognormal()
mod <- heter_model(b = 0, x = closed_form(~-ori_x))
ori_dat <- mod$gen(300)

@@

```{r over-under, fig.pos="!h", fig.cap = "Scatter plots for difference in $D$ and $\\hat{D}$ vs $\\sigma$ on test data for the $32 \\times 32$ optimized model. The data is grouped by whether the regression has only non-linearity violation, and whether it includes a second predictor in the regression formula. The brown lines are smoothing curves produced by fitting generalized additive models. The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.", dev = 'png', dpi = 300}
model_pred %>%
  filter(res == 32L) %>%
  filter(type == "test") %>%
  mutate(only_z = include_z & !include_heter & !include_non_normal) %>%
  rename(`Only has non-linearity violation` = only_z, `Has second predictor` = include_x2) %>%
  ggplot() +
  geom_hline(yintercept = 0, alpha = 0.5) +
  annotate("rect", 
           ymin = min(model_pred$effect_size[model_pred$type == "test"] - model_pred$vss[model_pred$type == "test"]),
                ymax = 0,
                xmin = min(model_pred$e_sigma),
                xmax = max(model_pred$e_sigma),
            fill = "#40B0A6",
            alpha = 0.3) +
  annotate("rect", 
           ymin = 0,
                ymax = max(model_pred$effect_size[model_pred$type == "test"] - model_pred$vss[model_pred$type == "test"]),
                xmin = min(model_pred$e_sigma),
                xmax = max(model_pred$e_sigma),
            fill = "#E1BE6A",
            alpha = 0.3) +
  geom_point(aes(e_sigma, effect_size - vss), alpha = 0.2) +
  geom_smooth(aes(e_sigma, effect_size - vss), se = FALSE, col = "#994F00") +
  facet_grid(`Has second predictor` ~ `Only has non-linearity violation`,
             labeller = label_both) +
  theme_light() +
  ylab(expression(D - hat(D))) +
  xlab(expression(sigma))
```

@@

Each lineup in @li2024plot contains one true residual plot and 19 null plots. While the true plot’s distance $D$ depends on the data-generating process, null plots have $D = 0$. We used our optimized computer vision model to estimate $D$ for all plots. To ensure fair comparison, we reject $H_0$ if the true plot has the highest estimated distance. Conventional tests, including the Ramsey RESET [@ramsey1969tests] for non-linearity and the Breusch-Pagan test [@breusch1979simple] for heteroskedasticity, were also applied to the same data.

Table \ref{tab:experiment-performance} reports performance metrics for $\hat{D}$ on true plots. While the performance is slightly worse than on the test data provided in Table \ref{tab:performance}, the mean absolute error remains low and the correlation between predictions and true values remains high. Consistent with results in Table \ref{tab:performance-sub}, non-linearity cases are harder to predict than heteroskedasticity cases.

@@

\noindent where $\bar{d}_{\text{true}}$ is the mean distance between the true plot and all null plots, and $\bar{d}_{\text{null}}^{(j)}$ is the mean distance from null plot $j$ to other nulls. These averages help address asymmetries in pairwise distances.

@@

Figure \ref{fig:delta} displays the scatter plot of the weighted detection rate against $\delta_{\text{adj}}$. It shows that weighted detection generally increases with $\delta_{\text{adj}}$, especially when it is positive. A negative $\delta_{\text{adj}}$ suggests some null plots may appear more visually distinctive to humans than the true plot. In some instances, the weighted detection rate is close to one, yet the $\delta_{\text{adj}}$ is negative. This discrepancy implies that the distance measure, or the estimated distance, may not perfectly reflect actual human behavior.

@@

```{r conv-mosaic, fig.pos = "!h", fig.cap = "Rejection rate ($p$-value $\\leq0.05$) of computer vision models conditional on conventional tests on non-linearity (left) and heteroskedasticity (right) lineups displayed using a mosaic plot. When the conventional test fails to reject, the computer vision mostly fails to reject the same plot as well as indicated by the height of the top right yellow rectangle, but there are non negliable amount of plots where the conventional test rejects but the computer vision model fails to reject as indicated by the width of the top left yellow rectangle.", fig.height = 4}
library(ggmosaic)

@@

```{r pcp, fig.cap = "Parallel coordinate plots of decisions made by computer vision model, conventional tests and visual tests made by human.", eval = TRUE, fig.height = 4}
library(ggpcp)
mutate(experiment, type = ifelse(type == "polynomial", "Non-linearity", "Heteroskedasticity")) %>%
  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity"))) %>%
  filter(!null_lineup, !attention_check) %>%
  select(type, actual_ss, model_reject_20, reject, conventional_reject) %>%
  mutate(test = rnorm(n())) %>%
  mutate(across(c(model_reject_20, conventional_reject, reject), ~ifelse(.x, "Reject", "Not reject"))) %>%
  mutate(agree = model_reject_20 == reject) %>%
  pcp_select(c(reject, model_reject_20, conventional_reject)) %>%
  group_by(type) %>%
  pcp_scale() %>%
  ungroup() %>%
  pcp_arrange(method = "from-right") %>%
  ggplot() +
  geom_pcp(aes_pcp(), linewidth = 0.1) +
  geom_pcp_labels(aes_pcp()) +
  geom_pcp_boxes(aes_pcp(), boxwidth = 0.1) +
  facet_wrap(~type, scales = "free_y") +
  theme_light() +
  scale_x_discrete(labels = c("Visual\ntest", "Computer\nvision\nmodel", "Conventional\ntest")) +
  labs(col = "D") +
  ylab("") +
  xlab("") +
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank(), 
        axis.title = element_blank())
```

```{r power, fig.pos = "!h", fig.cap = "Comparison of power of visual tests, conventional tests and the computer vision model. Marks along the x-axis at the bottom of the plot represent rejections made by each type of test. Marks at the top of the plot represent acceptances. Power curves are fitted by logistic regression models with no intercept but an offset equals to $\\text{log}(0.05/0.95)$.", fig.height = 4}

@@

```{r delta, fig.pos = "!h", fig.cap = "A weighted detection rate vs adjusted $\\delta$-difference plot. The brown line is smoothing curve produced by fitting generalized additive models.", fig.height = 4}
experiment %>%
  left_join(lineup_vss %>%
              group_by(unique_lineup_id) %>%
              summarise(delta_diff = first(delta_diff), gamma_diff = first(gamma_diff))) %>%
  ggplot() +
  geom_vline(xintercept = 0, alpha = 0.4, linetype = 2) +
  geom_point(aes(delta_diff, prop_detect), alpha = 0.3) +
  geom_smooth(aes(delta_diff, prop_detect), se = FALSE, col = "#994F00") +
  ylab("Weighted detection rate") +
  xlab(expression(delta[adj])) +
  theme_light()
```

@@

In this section, we present the performance of trained computer vision model on three example datasets. These include the dataset associated with the residual plot displaying a "left-triangle" shape, as displayed in Figure \ref{fig:false-finding}, along with the Boston housing dataset [@harrison1978hedonic], and the "dino" datasets from the `datasauRus` R package [@datasaurus].

The first example illustrates a scenario where both the computer vision model and human visual inspection successfully avoid rejecting $H_0$ when $H_0$ is true, contrary to conventional tests. This underscores the necessity of visually examining the residual plot.

In the second example, we encounter a more pronounced violation of the model, resulting in rejection of $H_0$ by all three tests. This highlights the practicality of the computer vision model, particularly for less intricate tasks.

The third example presents a situation where the model deviation is non-typical. Here, the computer vision model and human visual inspection reject $H_0$, whereas some commonly used conventional tests do not. This emphasizes the benefits of visual inspection and the unique advantage of the computer vision model, which, like humans, makes decisions based on visual discoveries.

@@

```{r false-lineup, fig.pos = "!h", fig.cap = 'A lineup of residual plots displaying "left-triangle" visual patterns. The true residual plot occupies position 10, yet there are no discernible visual patterns that distinguish it from the other plots.', fig.height = 8}
pos <- 10
result <- tibble()
for (i in c(1:20)[-pos]) {
  result <- bind_rows(result,
                      my_vi$rotate_resid() %>%
                        mutate(k = i))
}
result <- bind_rows(my_vi$get_fitted_and_resid() %>% 
                      mutate(k = pos),
                    result)
result %>%
  VI_MODEL$plot_lineup(remove_grid_line = TRUE,
                       remove_axis = TRUE,
                       theme = theme_light())
```

@@

Figure \ref{fig:boston-check} displays the residual plot and the assessment conducted by the computer vision model. A clear non-linearity pattern resembling a "U" shape is shown in the plot A. Furthermore, the RESET test yields a very small $p$-value. The estimated distance $\hat{D}$ significantly exceeds $Q_{null}(0.95)$, leading to rejection of $H_0$. The bootstrapped distribution also suggests that almost all the bootstrapped fitted models will be rejected, indicating that the fitted model is unlikely to be correctly specified. The attention map in plot B suggests the center of the image has higher leverage than other areas, and it is the turning point of the "U" shape. The PCA provided in plot D shows two distinct clusters of data points, further underling the visual differences between bootstrapped plots and null plots. This coincides the findings from Figure \ref{fig:boston-lineup}, where the true plot exhibiting a "U" shape is visually distinctive from null plots. If a visual test is conducted by human, $H_0$ will also be rejected.

@@

```{r boston-lineup, fig.pos = "!h", fig.cap = 'A lineup of residual plots for the Boston housing fitted model. The true residual plot is at position 7. It can be easily identified as the most different plot.', fig.height = 8}
pos <- 7
result <- tibble()
for (i in c(1:20)[-pos]) {
  result <- bind_rows(result,
                      my_vi$rotate_resid() %>%
                        mutate(k = i))
}
result <- bind_rows(my_vi$get_fitted_and_resid() %>% 
                      mutate(k = pos),
                    result)
result %>%
  VI_MODEL$plot_lineup(remove_grid_line = TRUE,
                       remove_axis = TRUE,
                       theme = theme_light())
```

@@

```{r dino-lineup, fig.pos = "!h", fig.cap = 'A lineup of residual plots for the fitted model on the "dinosaur" dataset. The true residual plot is at position 17. It can be easily identified as the most different plot as the visual pattern is extremely artificial.', fig.height = 8}
pos <- 17
result <- tibble()
for (i in c(1:20)[-pos]) {
  result <- bind_rows(result,
                      my_vi$rotate_resid() %>%
                        mutate(k = i))
}
result <- bind_rows(my_vi$get_fitted_and_resid() %>% 
                      mutate(k = pos),
                    result)
result %>%
  VI_MODEL$plot_lineup(remove_grid_line = TRUE,
                       remove_axis = TRUE,
                       theme = theme_light())
```

@@

The proposed distance measure assumes that the true model is a classical normal linear regression model, which can be restrictive. Although this paper does not address the relaxation of this assumption, there are potential methods to evaluate other types of regression models. The most comprehensive approach would be to define a distance measure for each different class of regression model and then train the computer vision model following the methodology described in this paper. To accelerate training, one could use the convolutional blocks of our trained model as a feature extractor and perform transfer learning on top of it, as these blocks effectively capture shapes in residual plots. Another approach would be to transform the residuals so they are roughly normally distributed and have constant variance. If only raw residuals are used, the distance-based statistical testing compares the difference in distance to a classical normal linear regression model for the true plot and null plots. This comparison is meaningful only if the difference can be identified by the distance measure proposed in this paper.

>>>>

```{r false-finding, fig.pos="!h", fig.width = 4, fig.height = 4*4/5, fig.pos="!h", fig.cap="An example residual vs fitted values plot (red line indicates 0 corresponds to the x-intercept, i.e. $y=0$). The vertical spread of the data points varies with the fitted values. This often indicates the existence of heteroskedasticity, however, here the result is due to skewed distribution of the predictors rather than heteroskedasticity. The Breusch-Pagan test rejects this residual plot at 95\\% significance level ($p\\text{-value} = 0.046$).", out.width = "50%"}
set.seed(452)
ori_x <- rand_lognormal()
mod <- heter_model(b = 0, x = closed_form(~-ori_x))
ori_dat <- mod$gen(300)

@@

```{r over-under, fig.pos="!h", fig.cap = "Scatter plots for difference in $D$ and $\\hat{D}$ vs $\\sigma$ on test data for the $32 \\times 32$ optimized model. The data is grouped by whether the regression has only non-linearity violation, and whether it includes a second predictor in the regression formula. The brown lines are smoothing curves produced by fitting generalized additive models. The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.", dev = 'png', dpi = 300, fig.height = 4.5}
model_pred %>%
  filter(res == 32L) %>%
  filter(type == "test") %>%
  mutate(only_z = include_z & !include_heter & !include_non_normal) %>%
  rename(`Only has non-linearity violation` = only_z, `Has second predictor` = include_x2) %>%
  ggplot() +
  geom_hline(yintercept = 0, alpha = 0.5) +
  annotate("rect", 
           ymin = min(model_pred$effect_size[model_pred$type == "test"] - model_pred$vss[model_pred$type == "test"]),
                ymax = 0,
                xmin = min(model_pred$e_sigma),
                xmax = max(model_pred$e_sigma),
            fill = "#40B0A6",
            alpha = 0.3) +
  annotate("rect", 
           ymin = 0,
                ymax = max(model_pred$effect_size[model_pred$type == "test"] - model_pred$vss[model_pred$type == "test"]),
                xmin = min(model_pred$e_sigma),
                xmax = max(model_pred$e_sigma),
            fill = "#E1BE6A",
            alpha = 0.3) +
  geom_point(aes(e_sigma, effect_size - vss), alpha = 0.2) +
  geom_smooth(aes(e_sigma, effect_size - vss), se = FALSE, col = "#994F00") +
  facet_grid(`Has second predictor` ~ `Only has non-linearity violation`,
             labeller = label_both) +
  theme_light() +
  ylab(expression(D - hat(D))) +
  xlab(expression(sigma))
```

@@

Each lineup in @li2024plot contains one true residual plot and 19 null plots. While the true residual plot’s distance $D$ depends on the data-generating process, null plots have $D = 0$. We used our optimized computer vision model to estimate $D$ for all plots. To ensure fair comparison, we reject $H_0$ if the true residual plot has the highest estimated distance. Conventional tests, including the Ramsey RESET [@ramsey1969tests] for non-linearity and the Breusch-Pagan test [@breusch1979simple] for heteroskedasticity, were also applied to the same data.

Table \ref{tab:experiment-performance} reports performance metrics for $\hat{D}$ on true residual plots. While the performance is slightly worse than on the test data provided in Table \ref{tab:performance}, the mean absolute error remains low and the correlation between predictions and true values remains high. Consistent with results in Table \ref{tab:performance-sub}, non-linearity cases are harder to predict than heteroskedasticity cases.

@@

\noindent where $\bar{d}_{\text{true}}$ is the mean distance between the true residual plot and all null plots, and $\bar{d}_{\text{null}}^{(j)}$ is the mean distance from null plot $j$ to other nulls. These averages help address asymmetries in pairwise distances.

@@

Figure \ref{fig:delta} displays the scatter plot of the weighted detection rate against $\delta_{\text{adj}}$. It shows that weighted detection generally increases with $\delta_{\text{adj}}$, especially when it is positive. This suggests that the differences in $\hat{D}$ capture some aspect of the visual separability between the true and null plots as perceived by humans, with larger positive differences generally indicating greater distinctiveness. However, the considerable variability in detection rates for similar $\delta_{\text{adj}}$ values implies that $\hat{D}$ does not fully account for all factors influencing human perception, and the alignment remains imperfect. A negative $\delta_{\text{adj}}$ suggests some null plots may appear more visually distinctive to humans than the true residual plot. In some instances, the weighted detection rate is close to one despite a negative $\delta_{\text{adj}}$. This discrepancy implies that the distance measure may not perfectly reflect actual human behavior.

@@

```{r conv-mosaic, fig.pos = "!h", fig.cap = "Rejection rate ($p$-value $\\leq0.05$) of computer vision models conditional on conventional tests on non-linearity (left) and heteroskedasticity (right) lineups displayed using a mosaic plot. When the conventional test fails to reject, the computer vision mostly fails to reject the same plot as well as indicated by the height of the top right yellow rectangle, but there are non negliable amount of plots where the conventional test rejects but the computer vision model fails to reject as indicated by the width of the top left yellow rectangle.", fig.height = 3}
library(ggmosaic)

@@

```{r pcp, fig.cap = "Parallel coordinate plots of decisions made by computer vision model, conventional tests and visual tests made by human.", eval = TRUE, fig.height = 3.5}
library(ggpcp)
mutate(experiment, type = ifelse(type == "polynomial", "Non-linearity", "Heteroskedasticity")) %>%
  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity"))) %>%
  filter(!null_lineup, !attention_check) %>%
  select(type, actual_ss, model_reject_20, reject, conventional_reject) %>%
  mutate(test = rnorm(n())) %>%
  mutate(across(c(model_reject_20, conventional_reject, reject), ~ifelse(.x, "Reject", "Not reject"))) %>%
  mutate(agree = model_reject_20 == reject) %>%
  pcp_select(c(reject, model_reject_20, conventional_reject)) %>%
  group_by(type) %>%
  pcp_scale() %>%
  ungroup() %>%
  pcp_arrange(method = "from-right") %>%
  ggplot() +
  geom_pcp(aes_pcp(), linewidth = 0.1) +
  geom_pcp_labels(aes_pcp()) +
  geom_pcp_boxes(aes_pcp(), boxwidth = 0.1) +
  facet_wrap(~type, scales = "free_y") +
  theme_light() +
  scale_x_discrete(labels = c("Visual\ntest", "Computer\nvision\nmodel", "Conventional\ntest")) +
  labs(col = "D") +
  ylab("") +
  xlab("") +
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank(), 
        axis.title = element_blank())
```

```{r power, fig.pos = "!h", fig.cap = "Comparison of power of visual tests, conventional tests and the computer vision model. Marks along the x-axis at the bottom of the plot represent rejections made by each type of test. Marks at the top of the plot represent acceptances. Power curves are fitted by logistic regression models with no intercept but an offset equals to $\\text{log}(0.05/0.95)$.", fig.width = 7, fig.height = 3.5}

@@

```{r delta, fig.pos = "!h", fig.cap = "A weighted detection rate vs adjusted $\\delta$-difference plot. The brown line is smoothing curve produced by fitting generalized additive models.", fig.width = 6, fig.height = 3, out.width = "80%"}
experiment %>%
  left_join(lineup_vss %>%
              group_by(unique_lineup_id) %>%
              summarise(delta_diff = first(delta_diff), gamma_diff = first(gamma_diff))) %>%
  ggplot() +
  geom_vline(xintercept = 0, alpha = 0.4, linetype = 2) +
  geom_point(aes(delta_diff, prop_detect), alpha = 0.3) +
  geom_smooth(aes(delta_diff, prop_detect), se = FALSE, col = "#994F00") +
  ylab("Weighted detection rate") +
  xlab(expression(delta[adj])) +
  theme_light()
```

@@

In this section, we present the performance of the trained computer vision model on three example datasets, including the dataset associated with the residual plot displaying a "left-triangle" shape (Figure \ref{fig:false-finding}), the Boston housing dataset [@harrison1978hedonic], and the "dino" datasets from the from the `datasauRus` package [@datasaurus].

In the first case, both the model and human inspection correctly avoid rejecting $H_0$ when it is true, contrary to conventional tests, highlighting the necessity of visual examination. The second case shows a clear model violation detected by all tests, demonstrating the model’s practical use for less intricate tasks. The third case involves an unusual deviation where the model and humans reject $H_0$, but some conventional tests do not, illustrating the benefits of visual-based decision-making.

@@

```{r false-lineup, fig.pos = "!h", fig.cap = 'A lineup of residual plots displaying "left-triangle" visual patterns. The true residual plot occupies position 10, yet there are no discernible visual patterns that distinguish it from the other plots.', fig.height = 8, out.width = "70%"}
pos <- 10
result <- tibble()
for (i in c(1:20)[-pos]) {
  result <- bind_rows(result,
                      my_vi$rotate_resid() %>%
                        mutate(k = i))
}
result <- bind_rows(my_vi$get_fitted_and_resid() %>% 
                      mutate(k = pos),
                    result)
result %>%
  VI_MODEL$plot_lineup(remove_grid_line = TRUE,
                       remove_axis = TRUE,
                       theme = theme_light())
```

@@

Figure \ref{fig:boston-check} displays the residual plot and the assessment conducted by the computer vision model. A clear non-linearity pattern resembling a "U" shape is shown in the plot A. Furthermore, the RESET test yields a very small $p$-value. The estimated distance $\hat{D}$ significantly exceeds $Q_{null}(0.95)$, leading to rejection of $H_0$. The bootstrapped distribution also suggests that almost all the bootstrapped fitted models will be rejected, indicating that the fitted model is unlikely to be correctly specified. The attention map in plot B suggests the center of the image has higher leverage than other areas, and it is the turning point of the "U" shape. The PCA provided in plot D shows two distinct clusters of data points, further underling the visual differences between bootstrapped plots and null plots. This coincides the findings from Figure \ref{fig:boston-lineup}, where the true residual plot exhibiting a "U" shape is visually distinctive from null plots. If a visual test is conducted by human, $H_0$ will also be rejected.

@@

```{r boston-lineup, fig.pos = "!h", fig.cap = 'A lineup of residual plots for the Boston housing fitted model. The true residual plot is at position 7. It can be easily identified as the most different plot.', fig.height = 8, out.width = "70%"}
pos <- 7
result <- tibble()
for (i in c(1:20)[-pos]) {
  result <- bind_rows(result,
                      my_vi$rotate_resid() %>%
                        mutate(k = i))
}
result <- bind_rows(my_vi$get_fitted_and_resid() %>% 
                      mutate(k = pos),
                    result)
result %>%
  VI_MODEL$plot_lineup(remove_grid_line = TRUE,
                       remove_axis = TRUE,
                       theme = theme_light())
```

@@

```{r dino-lineup, fig.pos = "!h", fig.cap = 'A lineup of residual plots for the fitted model on the "dinosaur" dataset. The true residual plot is at position 17. It can be easily identified as the most different plot as the visual pattern is extremely artificial.', fig.height = 8, out.width = "70%"}
pos <- 17
result <- tibble()
for (i in c(1:20)[-pos]) {
  result <- bind_rows(result,
                      my_vi$rotate_resid() %>%
                        mutate(k = i))
}
result <- bind_rows(my_vi$get_fitted_and_resid() %>% 
                      mutate(k = pos),
                    result)
result %>%
  VI_MODEL$plot_lineup(remove_grid_line = TRUE,
                       remove_axis = TRUE,
                       theme = theme_light())
```

@@

The proposed distance measure assumes that the true model is a classical normal linear regression model, which can be restrictive. Although this paper does not address the relaxation of this assumption, there are potential methods to evaluate other types of regression models. The most comprehensive approach would be to define a distance measure for each different class of regression model and then train the computer vision model following the methodology described in this paper. To accelerate training, one could use the convolutional blocks of our trained model as a feature extractor and perform transfer learning on top of it, as these blocks effectively capture shapes in residual plots. Another approach would be to transform the residuals so they are roughly normally distributed and have constant variance. If only raw residuals are used, the distance-based statistical testing compares the difference in distance to a classical normal linear regression model for the true residual plot and null plots. This comparison is meaningful only if the difference can be identified by the distance measure proposed in this paper.

### 2025-06-16 paper.tex
Adjust plot size

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/false-finding-1}

@@

Each lineup in \citet{li2024plot} contains one true residual plot and 19
null plots. While the true plot's distance \(D\) depends on the
data-generating process, null plots have \(D = 0\). We used our
optimized computer vision model to estimate \(D\) for all plots. To
ensure fair comparison, we reject \(H_0\) if the true plot has the
highest estimated distance. Conventional tests, including the Ramsey
RESET \citep{ramsey1969tests} for non-linearity and the Breusch-Pagan
test \citep{breusch1979simple} for heteroskedasticity, were also applied
to the same data.

Table \ref{tab:experiment-performance} reports performance metrics for
\(\hat{D}\) on true plots. While the performance is slightly worse than
on the test data provided in Table \ref{tab:performance}, the mean
absolute error remains low and the correlation between predictions and
true values remains high. Consistent with results in Table
\ref{tab:performance-sub}, non-linearity cases are harder to predict
than heteroskedasticity cases.

@@

\noindent where \(\bar{d}_{\text{true}}\) is the mean distance between
the true plot and all null plots, and \(\bar{d}_{\text{null}}^{(j)}\) is
the mean distance from null plot \(j\) to other nulls. These averages
help address asymmetries in pairwise distances.

@@

Figure \ref{fig:delta} displays the scatter plot of the weighted
detection rate against \(\delta_{\text{adj}}\). It shows that weighted
detection generally increases with \(\delta_{\text{adj}}\), especially
when it is positive. A negative \(\delta_{\text{adj}}\) suggests some
null plots may appear more visually distinctive to humans than the true
plot. In some instances, the weighted detection rate is close to one,
yet the \(\delta_{\text{adj}}\) is negative. This discrepancy implies
that the distance measure, or the estimated distance, may not perfectly
reflect actual human behavior.

@@

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/delta-1}

@@

In this section, we present the performance of trained computer vision
model on three example datasets. These include the dataset associated
with the residual plot displaying a ``left-triangle'' shape, as
displayed in Figure \ref{fig:false-finding}, along with the Boston
housing dataset \citep{harrison1978hedonic}, and the ``dino'' datasets
from the \texttt{datasauRus} R package \citep{datasaurus}.

The first example illustrates a scenario where both the computer vision
model and human visual inspection successfully avoid rejecting \(H_0\)
when \(H_0\) is true, contrary to conventional tests. This underscores
the necessity of visually examining the residual plot.

In the second example, we encounter a more pronounced violation of the
model, resulting in rejection of \(H_0\) by all three tests. This
highlights the practicality of the computer vision model, particularly
for less intricate tasks.

The third example presents a situation where the model deviation is
non-typical. Here, the computer vision model and human visual inspection
reject \(H_0\), whereas some commonly used conventional tests do not.
This emphasizes the benefits of visual inspection and the unique
advantage of the computer vision model, which, like humans, makes
decisions based on visual discoveries.

@@

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/false-lineup-1}

@@

Figure \ref{fig:boston-check} displays the residual plot and the
assessment conducted by the computer vision model. A clear non-linearity
pattern resembling a ``U'' shape is shown in the plot A. Furthermore,
the RESET test yields a very small \(p\)-value. The estimated distance
\(\hat{D}\) significantly exceeds \(Q_{null}(0.95)\), leading to
rejection of \(H_0\). The bootstrapped distribution also suggests that
almost all the bootstrapped fitted models will be rejected, indicating
that the fitted model is unlikely to be correctly specified. The
attention map in plot B suggests the center of the image has higher
leverage than other areas, and it is the turning point of the ``U''
shape. The PCA provided in plot D shows two distinct clusters of data
points, further underling the visual differences between bootstrapped
plots and null plots. This coincides the findings from Figure
\ref{fig:boston-lineup}, where the true plot exhibiting a ``U'' shape is
visually distinctive from null plots. If a visual test is conducted by
human, \(H_0\) will also be rejected.

@@

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/boston-lineup-1}

@@

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/dino-lineup-1}

>>>>

{\centering \includegraphics[width=0.5\linewidth]{paper_files/figure-latex/false-finding-1}

@@

Each lineup in \citet{li2024plot} contains one true residual plot and 19
null plots. While the true residual plot's distance \(D\) depends on the
data-generating process, null plots have \(D = 0\). We used our
optimized computer vision model to estimate \(D\) for all plots. To
ensure fair comparison, we reject \(H_0\) if the true residual plot has
the highest estimated distance. Conventional tests, including the Ramsey
RESET \citep{ramsey1969tests} for non-linearity and the Breusch-Pagan
test \citep{breusch1979simple} for heteroskedasticity, were also applied
to the same data.

Table \ref{tab:experiment-performance} reports performance metrics for
\(\hat{D}\) on true residual plots. While the performance is slightly
worse than on the test data provided in Table \ref{tab:performance}, the
mean absolute error remains low and the correlation between predictions
and true values remains high. Consistent with results in Table
\ref{tab:performance-sub}, non-linearity cases are harder to predict
than heteroskedasticity cases.

@@

\noindent where \(\bar{d}_{\text{true}}\) is the mean distance between
the true residual plot and all null plots, and
\(\bar{d}_{\text{null}}^{(j)}\) is the mean distance from null plot
\(j\) to other nulls. These averages help address asymmetries in
pairwise distances.

@@

Figure \ref{fig:delta} displays the scatter plot of the weighted
detection rate against \(\delta_{\text{adj}}\). It shows that weighted
detection generally increases with \(\delta_{\text{adj}}\), especially
when it is positive. This suggests that the differences in \(\hat{D}\)
capture some aspect of the visual separability between the true and null
plots as perceived by humans, with larger positive differences generally
indicating greater distinctiveness. However, the considerable
variability in detection rates for similar \(\delta_{\text{adj}}\)
values implies that \(\hat{D}\) does not fully account for all factors
influencing human perception, and the alignment remains imperfect. A
negative \(\delta_{\text{adj}}\) suggests some null plots may appear
more visually distinctive to humans than the true residual plot. In some
instances, the weighted detection rate is close to one despite a
negative \(\delta_{\text{adj}}\). This discrepancy implies that the
distance measure may not perfectly reflect actual human behavior.

@@

{\centering \includegraphics[width=0.8\linewidth]{paper_files/figure-latex/delta-1}

@@

In this section, we present the performance of the trained computer
vision model on three example datasets, including the dataset associated
with the residual plot displaying a ``left-triangle'' shape (Figure
\ref{fig:false-finding}), the Boston housing dataset
\citep{harrison1978hedonic}, and the ``dino'' datasets from the from the
\texttt{datasauRus} package \citep{datasaurus}.

In the first case, both the model and human inspection correctly avoid
rejecting \(H_0\) when it is true, contrary to conventional tests,
highlighting the necessity of visual examination. The second case shows
a clear model violation detected by all tests, demonstrating the model's
practical use for less intricate tasks. The third case involves an
unusual deviation where the model and humans reject \(H_0\), but some
conventional tests do not, illustrating the benefits of visual-based
decision-making.

@@

{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/false-lineup-1}

@@

Figure \ref{fig:boston-check} displays the residual plot and the
assessment conducted by the computer vision model. A clear non-linearity
pattern resembling a ``U'' shape is shown in the plot A. Furthermore,
the RESET test yields a very small \(p\)-value. The estimated distance
\(\hat{D}\) significantly exceeds \(Q_{null}(0.95)\), leading to
rejection of \(H_0\). The bootstrapped distribution also suggests that
almost all the bootstrapped fitted models will be rejected, indicating
that the fitted model is unlikely to be correctly specified. The
attention map in plot B suggests the center of the image has higher
leverage than other areas, and it is the turning point of the ``U''
shape. The PCA provided in plot D shows two distinct clusters of data
points, further underling the visual differences between bootstrapped
plots and null plots. This coincides the findings from Figure
\ref{fig:boston-lineup}, where the true residual plot exhibiting a ``U''
shape is visually distinctive from null plots. If a visual test is
conducted by human, \(H_0\) will also be rejected.

@@

{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/boston-lineup-1}

@@

{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/dino-lineup-1}
